---
title: "Capstone_Project_Data_Manipulation"
author: "Cathy Kiriakos"
date: "September 12, 2017"
output: html_document
---

```{r setwd, include=FALSE}
path <- file.path("C:", "Users","kiriakosc", ".git", "Springboard_Capstone")
setwd(path)
```

```{r list_wd_files, include=FALSE}
list.files(path)
```


```{r load_csvs, warning=FALSE, include=FALSE}
Capstone_Avantis_Transformed <-read.csv(file.path(path,"Capstone_Avantis_Transformed.csv")) #Avantis base data
Capstone_GL_Transformed <-read.csv(file.path(path,"Capstone_GL_Transformed.csv")) #GL base data
Budget_Information <- read.csv(file.path(path,"Budget Information.csv"))
```

#Load needed packages
```{r load_packages, include=FALSE}
library(magrittr)
library(formatR)
library(dplyr)
library(tidyr)
library(devtools)
```
#Transform Avantis Data: 
Step 1 pull apart GL data using tidyr: note cntrl-shift-m = pipe operator shortcut, place "" around column names!
```{r Transform_Avantis_Data}
Avantis_Transformed <- Capstone_Avantis_Transformed %>% 
  separate("G.L.Account.number", c("Company Id","Account Number", "Expense Code", "Department","Location","Activity Code","Center" ), sep = "\\-") %>% 
  #Transform Avantis Data: Step 2 pull apart date into month and year
  separate("Processing.Date", c("Month","Day", "Year" ), sep = "\\/")
#Trim leading and trailing white space from columns
Avantis_Transformed <- data.frame(lapply(Avantis_Transformed,trimws), stringsAsFactors = FALSE)

#View Output  
head(Avantis_Transformed,10)
```
#Transform GL Data: 
Step 2: break apart GL Description to identifiers - Segment: (INE or ISS, all others will remain whole), Recevable/Non Recevable/Cat Name (1000, 1001, 1010) or if issue catalog #, PO Number, Invoice Number/Warehouse Detail -  separate gl desc
```{r Transform_GL_Data}
GL_Transformed <- Capstone_GL_Transformed %>% 
  separate("descp", c("Segment","Recevable/Catalog Code", "PO Number", "Invoice Number/Warehouse Detail" ), sep = "\\-", fill = "left") 
#Trim leading and trailing white space from columns
GL_Transformed <- data.frame(lapply(GL_Transformed,trimws), stringsAsFactors = FALSE)

head(GL_Transformed,10)
```
Test Vector Types 
```{r Test_Vector_Types, include=FALSE}
#Test vector Types 
is.numeric(GL_Transformed$account)
is.numeric(GL_Transformed$expense_code)
is.numeric(GL_Transformed$department)
is.numeric(GL_Transformed$location)
is.numeric(GL_Transformed$activity)
is.numeric(GL_Transformed$SumOftrans_amt)
```
View Data frame to ensure merge will funciton
```{r View_Data_Avantis_Tranformed,include=FALSE}
head(Avantis_Transformed)
```
View Data frame to ensure merge will funciton
```{r View_GL_Transformed,include=FALSE}
head(GL_Transformed)
str(GL_Transformed)
```

```{r Check_Vector_Types,include=FALSE}
#check vector type
typeof(GL_Transformed$account)
typeof(Avantis_Transformed$Account.Number)
typeof(GL_Transformed$department)
typeof(Avantis_Transformed$Department)
typeof(GL_Transformed$expense_code)
typeof(Avantis_Transformed$Expense.Code)
typeof(GL_Transformed$posting_yr)
typeof(GL_Transformed$posting_pd)
typeof(Avantis_Transformed$Month)
```
Merge GL with Avantis Data
```{r MergeGL&Avantis}
GL_Final <-  merge( GL_Transformed,Avantis_Transformed, 
                                        by.x = c("account","department","expense_code", "SumOftrans_amt","posting_yr","posting_pd"),
                                        by.y = c("Account.Number","Department","Expense.Code","Extended.line.amount_1","Year","Month"),all.x = TRUE)
```
Break Apart Code Lookup to make merge for descriptions function
```{r Break_Apart_Code_Lookup}
Activity_Code_Lookup <- select(Code_Lookup, Activity.Code, Activity.Description)
Expense_Code_Lookup <- select(Code_Lookup, Expense.Code, Expense.Code.Description, Expense.Code.Category)
Account_Code_Lookup <- select(Code_Lookup, Account.Number, Account.Description)
```
Convert Account Code Lookups to characer for merge
```{r Load_Account_Lookup_ConvertAsCharacter}
Code_Lookup <-read.csv(file.path(path,"Code Lookup.csv"), header = TRUE, stringsAsFactors = FALSE)

Code_Lookup[,"Expense.Code"] <- as.character(Code_Lookup[,"Expense.Code"])
Code_Lookup[,"Activity.Code"] <- as.character(Code_Lookup[,"Activity.Code"])
Code_Lookup[,"Account.Number"] <- as.character(Code_Lookup[,"Account.Number"])

Code_Lookup <- data.frame(lapply(Code_Lookup,trimws), stringsAsFactors = FALSE)
str(Code_Lookup)
```
Merge GL_Final with Activity Code Descriptions
```{r Merge_GL_Final_AccountLookups, error=FALSE}
GL_Final <- merge(GL_Final, Activity_Code_Lookup,
                  by.x = c("activity"),
                  by.y = c("Activity.Code"), all.x = TRUE)
```
Merge GL_Final with Expense Code Descriptions
```{r}
GL_Final <- merge(GL_Final, Expense_Code_Lookup,
                  by.x = c("expense_code"),
                  by.y = c("Expense.Code"), all.x = TRUE)
```
Merge GL_Final with Account Code Descriptions 
```{r}
GL_Final <- merge(GL_Final, Account_Code_Lookup,
                  by.x = c("account"),
                  by.y = c("Account.Number"), all.x = TRUE)
```
Prepare Budget data:
```{r Prepare_BudgetData}
# 1) Break apart F9 Budget data 
Budget_Information_Transformed <- Budget_Information %>% 
  separate("Account", c("Company_Code", "Account_Number", "Expense_Code", "Department", "Location", "Activity_Code","Center"), sep = "\\-", fill = "left")
#2) Trim Whitespace
Budget_Information_Transformed <- data.frame(lapply(Budget_Information_Transformed, trimws), stringsAsFactors = FALSE)
```
Bind budget to gl data:
```{r Bind_GL&Budget}
GL_Final <- bind_rows(GL_Final, Budget_Information_Transformed)
```
Convert SumofTrans to numeric
```{r Convert_SumofTranstoNumer, error=FALSE}
#Convert Sum of Trans Amount to Numeric for analysis
GL_Final[,"SumOftrans_amt"] <- as.numeric(GL_Final[,"SumOftrans_amt"])
```
Convert na's to blank
```{r, nastoblank}
GL_Final[is.na(GL_Final)] <- ""
```

Incorporate and clean up Budget Data in GL 
```{r Incorporate_BudgetData}
#Incorporate Budget into base data 
GL_Final$account <- ifelse(GL_Final$account == "", GL_Final$Account_Number, GL_Final$account)
GL_Final$expense_code <- ifelse(GL_Final$expense_code == "", GL_Final$Expense_Code, GL_Final$expense_code)
GL_Final$activity <- ifelse(GL_Final$activity == "", GL_Final$Activity_Code, GL_Final$activity)
GL_Final$department <- ifelse(GL_Final$department == "", GL_Final$Department, GL_Final$department)
GL_Final$SumOftrans_amt <- ifelse(GL_Final$SumOftrans_amt == "", GL_Final$Amount, GL_Final$SumOftrans_amt)
GL_Final$posting_yr <- ifelse(GL_Final$posting_yr == "", GL_Final$Years, GL_Final$posting_yr)
GL_Final$posting_pd <- ifelse(GL_Final$posting_pd == "", GL_Final$Period, GL_Final$posting_pd) 
GL_Final$location <- ifelse(GL_Final$location == "", GL_Final$Location, GL_Final$location)
GL_Final$Types <- ifelse(GL_Final$Types == "","Actual",GL_Final$Types)
```
Consolidate data & remove unneeded columns
```{r Consolidate_data}
GL_Final$Years  = NULL
GL_Final$Currencies = NULL
GL_Final$Company_Code = NULL
GL_Final$Account_Number = NULL
GL_Final$Expense_Code = NULL
GL_Final$Department= NULL
GL_Final$Activity_Code = NULL
GL_Final$Description = NULL
GL_Final$Period = NULL
GL_Final$Amount = NULL
GL_Final$Companies = NULL

head(GL_Final)
```
Replace GL_Final NAs with blank
```{r Replace_GL_NAs}
GL_Final[is.na(GL_Final)] <- ""
```

View Stucture:
```{r}
str(GL_Final)
```
Change posting_period  and account number to numeric for season classification and account designation
```{r, post_per_numeric}
GL_Final[,"posting_pd"] <- as.numeric(GL_Final[,"posting_pd"])
GL_Final[,"account"] <- as.numeric(GL_Final[,"account"])
```
View Head
```{r,head}
head(GL_Final)
```

Add Capital or O&M designation GL_Final
```{r Add_AccountDesignations}
GL_Final$Designation <- ""
GL_Final$Designation <- ifelse(GL_Final$account < 140000,"Capital" ," O&M")
```

Incorporate season into data for Time Series Analysis: 10/25 4:27 only returning spring
```{r, season,error=FALSE}
GL_Final$Season <- ifelse(GL_Final$posting_pd == c(12,1,2), "Winter",
                    ifelse(GL_Final$posting_pd == c(3,4,5),"Spring",
                    ifelse(GL_Final$posting_pd == c(6,7,8), "Summer",
                    ifelse(GL_Final$posting_pd == c(9,10,11),"fall",""))))
```





