---
title: "Springboard_Capstone_Stats"
author: "Cathy Kiriakos"
date: "November 17, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Choose appropriate libraries for stats calcs
```{r, include=FALSE}
library(reshape)
library(reshape2)
library(tables)
library(ggplot2)
```
Set options
```{r, include = FALSE}
options(scipen = 999)
options(digits = 2)
```
Load File: 
```{r, load}
path <- file.path("C:", "Users","kiriakosc", ".git", "Springboard_Capstone") 
GL_Final <-read.csv(file.path(path,"GL_Final.csv"),header = TRUE, stringsAsFactors = FALSE ) 
```


##Can you count something interesting? 
Counts in budget data are not very indiciative of trends, but sums by category are.  Below we will see summaries of O&M expenditures by year.

The first step is to subset data by O&M budget & expeditures by year: 
```{r,results_by_yr}
Results_2015 <- filter(GL_Final,grepl("2015", posting_yr) &  grepl("O&M", Designation) &  grepl("ACTUAL",Types))
Budget_2015 <- filter(GL_Final,grepl("2015", posting_yr) &  grepl("O&M", Designation) &  grepl("BUDGET",Types))

head(Results_2015)
head(Budget_2015)

Results_2016 <- filter(GL_Final,grepl("2016", posting_yr) &  grepl("O&M", Designation) &  grepl("ACTUAL",Types))
Budget_2016 <- filter(GL_Final,grepl("2016", posting_yr) &  grepl("O&M", Designation) &  grepl("BUDGET",Types))

head(Results_2016)
head(Budget_2016)

Results_2017 <- filter(GL_Final,grepl("2017", posting_yr) &  grepl("O&M", Designation) &  grepl("ACTUAL",Types))
Budget_2017 <- filter(GL_Final,grepl("2017", posting_yr) &  grepl("O&M", Designation) &  grepl("BUDGET",Types))

head(Results_2017)
head(Budget_2017)
```
Next we will create tables to view expenditures by expense code, activity code and account number within their respective years vs. their respected budget amounts:
Starting with a 2015 summary: here we will melt the data to indicate id variables & measure variables 
```{r, melt}
Melt2015Res <- melt(Results_2015, id.vars = c("Expense.Code.Description", "Activity.Description", "account", "Types", "Season"),measure.vars = "SumOftrans_amt", na.rm = TRUE)

Melt2015Bud <- melt(Budget_2015, id.vars = c("Expense.Code.Description", "Activity.Description", "account", "Types", "Season"),measure.vars = "SumOftrans_amt", na.rm = TRUE)

Melt2016Res <- melt(Results_2016, id.vars = c("Expense.Code.Description", "Activity.Description", "account", "Types", "Season"),measure.vars = "SumOftrans_amt", na.rm = TRUE)

Melt2016Bud <- melt(Budget_2016, id.vars = c("Expense.Code.Description", "Activity.Description", "account", "Types", "Season"),measure.vars = "SumOftrans_amt", na.rm = TRUE)

Melt2017Res <- melt(Results_2017, id.vars = c("Expense.Code.Description", "Activity.Description", "account", "Types", "Season"),measure.vars = "SumOftrans_amt", na.rm = TRUE)

Melt2017Bud <- melt(Budget_2017, id.vars = c("Expense.Code.Description", "Activity.Description", "account", "Types", "Season"),measure.vars = "SumOftrans_amt", na.rm = TRUE)

#view first few rows of melted data sets
head(Melt2015Res)
head(Melt2015Bud)
head(Melt2017Res)
head(Melt2017Bud)
head(Melt2017Res)
head(Melt2017Bud)
```
Now that the data is melted we can cast to build a pivot table summary, here we can see expenditure totals for each season in a given year: 

```{r, cast}
cast(Melt2015Res, Expense.Code.Description + Activity.Description + account + Types + Season ~ variable, sum, margins = TRUE)

cast(Melt2015Bud, Expense.Code.Description + Activity.Description + account + Types + Season ~ variable, sum, margins = TRUE)

cast(Melt2016Res, Expense.Code.Description + Activity.Description + account + Types + Season ~ variable, sum, margins = TRUE)

cast(Melt2016Bud, Expense.Code.Description + Activity.Description + account + Types + Season ~ variable, sum, margins = TRUE)

cast(Melt2017Res, Expense.Code.Description + Activity.Description + account + Types + Season ~ variable, sum, margins = TRUE)

cast(Melt2017Bud, Expense.Code.Description + Activity.Description + account + Types + Season ~ variable, sum, margins = TRUE)

```

##Can you find some trends (high, low, increase, decrease, anomalies)?
Contracted Services & Materials/Supplies - expense codes 62 and 47 respectively are the categories that vary exponentially, and would be the best to look at, we will first start by subsetting the budget and expenditure data for the expense code and season below: 
```{r, subsetdata}
#2015 EC 62 Subsetted data by season:

CS15SPRES <- subset(Melt2015Res,Season = "Spring", expense_code = "62")
CS15SPBUD <- subset(Melt2015Bud, Season = "Spring", expense_code = "62")

CS15SURES <- subset(Melt2015Res,Season = "Summer", expense_code = "62")
CS15SUBUD <- subset(Melt2015Bud, Season = "Summer", expense_code = "62")

CS15WIRES <- subset(Melt2015Res,Season = "Winter", expense_code = "62")
CS15WIBUD <- subset(Melt2015Bud, Season = "Winter", expense_code = "62")

CS15FARES <- subset(Melt2015Res,Season = "Fall", expense_code = "62")
CS15FABUD <- subset(Melt2015Bud, Season = "Fall", expense_code = "62")

#2017 EC 62 Subsetted data by season:

CS16SPRES <- subset(Melt2016Res,Season = "Spring", expense_code = "62")
CS16SPBUD <- subset(Melt2016Bud, Season = "Spring", expense_code = "62")

CS16SURES <- subset(Melt2016Res,Season = "Summer", expense_code = "62")
CS16SUBUD <- subset(Melt2016Bud, Season = "Summer", expense_code = "62")

CS16WIRES <- subset(Melt2016Res,Season = "Winter", expense_code = "62")
CS16WIBUD <- subset(Melt2016Bud, Season = "Winter", expense_code = "62")

CS16FARES <- subset(Melt2016Res,Season = "Fall", expense_code = "62")
CS16FABUD <- subset(Melt2016Bud, Season = "Fall", expense_code = "62")

#2017 EC 62 Subsetted data by season:

CS17SPRES <- subset(Melt2017Res,Season = "Spring", expense_code = "62")
CS17SPBUD <- subset(Melt2017Bud, Season = "Spring", expense_code = "62")

CS17SURES <- subset(Melt2017Res,Season = "Summer", expense_code = "62")
CS17SUBUD <- subset(Melt2017Bud, Season = "Summer", expense_code = "62")

CS17WIRES <- subset(Melt2017Res,Season = "Winter", expense_code = "62")
CS17WIBUD <- subset(Melt2017Bud, Season = "Winter", expense_code = "62")

CS17FARES <- subset(Melt2017Res,Season = "Fall", expense_code = "62")
CS17FABUD <- subset(Melt2017Bud, Season = "Fall", expense_code = "62")

#2015 EC 47 Subsetted data by season:

MS15SPRES <- subset(Melt2015Res,Season = "Spring", expense_code = "47")
MS15SPBUD <- subset(Melt2015Bud, Season = "Spring", expense_code = "47")

MS15SURES <- subset(Melt2015Res,Season = "Summer", expense_code = "47")
MS15SUBUD <- subset(Melt2015Bud, Season = "Summer", expense_code = "47")

MS15WIRES <- subset(Melt2015Res,Season = "Winter", expense_code = "47")
MS15WIBUD <- subset(Melt2015Bud, Season = "Winter", expense_code = "47")

MS15FARES <- subset(Melt2015Res,Season = "Fall", expense_code = "47")
MS15FABUD <- subset(Melt2015Bud, Season = "Fall", expense_code = "47")

#2016 EC 47 Subsetted data by season:

MS16SPRES <- subset(Melt2016Res,Season = "Spring", expense_code = "47")
MS16SPBUD <- subset(Melt2016Bud, Season = "Spring", expense_code = "47")

MS16SURES <- subset(Melt2016Res,Season = "Summer", expense_code = "47")
MS16SUBUD <- subset(Melt2016Bud, Season = "Summer", expense_code = "47")

MS16WIRES <- subset(Melt2016Res,Season = "Winter", expense_code = "47")
MS16WIBUD <- subset(Melt2016Bud, Season = "Winter", expense_code = "47")

MS16FARES <- subset(Melt2016Res,Season = "Fall", expense_code = "47")
MS16FABUD <- subset(Melt2016Bud, Season = "Fall", expense_code = "47")

#2017 EC 47 Subsetted data by season:

MS17SPRES <- subset(Melt2017Res,Season = "Spring", expense_code = "47")
MS17SPBUD <- subset(Melt2017Bud, Season = "Spring", expense_code = "47")

MS17SURES <- subset(Melt2017Res,Season = "Summer", expense_code = "47")
MS17SUBUD <- subset(Melt2017Bud, Season = "Summer", expense_code = "47")

MS17WIRES <- subset(Melt2017Res,Season = "Winter", expense_code = "47")
MS17WIBUD <- subset(Melt2017Bud, Season = "Winter", expense_code = "47")

MS17FARES <- subset(Melt2017Res,Season = "Fall", expense_code = "47")
MS17FABUD <- subset(Melt2017Bud, Season = "Fall", expense_code = "47")
```
Now that the data is subsetted we can look at mean, min, and max for each category as compared to its budget 
```{r}
#2015 Contracted Services Actuals vs. Budget: 
#Spring
format(cast(CS15SPRES, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

format(cast(CS15SPBUD, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

#Summer
format(cast(CS15SURES, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

format(cast(CS15SUBUD, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

#Fall
format(cast(CS15FARES, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

format(cast(CS15FABUD, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

#Winter:
format(cast(CS15WIRES, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

format(cast(CS15WIBUD, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)
```
```{r}
#2016 Contracted Services Actuals vs. Budget: 
#Spring
format(cast(CS16SPRES, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

format(cast(CS16SPBUD, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

#Summer
format(cast(CS16SURES, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

format(cast(CS16SUBUD, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

#Fall
format(cast(CS16FARES, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

format(cast(CS16FABUD, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

#Winter:
format(cast(CS16WIRES, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

format(cast(CS16WIBUD, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)
```


```{r}
#2017 Contracted Services Actuals vs. Budget: 
#Spring
format(cast(CS17SPRES, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

format(cast(CS17SPBUD, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

#Summer
format(cast(CS17SURES, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

format(cast(CS17SUBUD, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

#Fall
format(cast(CS17FARES, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

format(cast(CS17FABUD, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

#Winter:
format(cast(CS17WIRES, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

format(cast(CS17WIBUD, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)
```
```{r}
#2015 Materials & Services Actuals vs. Budget: 
#Spring
format(cast(MS15SPRES, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

format(cast(MS15SPBUD, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

#Summer
format(cast(MS15SURES, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

format(cast(MS15SUBUD, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

#Fall
format(cast(MS15FARES, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

format(cast(MS15FABUD, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

#Winter:
format(cast(MS15WIRES, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

format(cast(MS15WIBUD, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)
```
```{r}
#2016 Materials & Services Actuals vs. Budget: 
#Spring
format(cast(MS16SPRES, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

format(cast(MS16SPBUD, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

#Summer
format(cast(MS16SURES, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

format(cast(MS16SUBUD, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

#Fall
format(cast(MS16FARES, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

format(cast(MS16FABUD, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

#Winter:
format(cast(MS16WIRES, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

format(cast(MS16WIBUD, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)
```


```{r}
#2017 Materials & Services Actuals vs. Budget: 
#Spring
format(cast(MS17SPRES, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

format(cast(MS17SPBUD, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

#Summer
format(cast(MS17SURES, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

format(cast(MS17SUBUD, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

#Fall
format(cast(MS17FARES, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

format(cast(MS17FABUD, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

#Winter:
format(cast(MS17WIRES, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)

format(cast(MS17WIBUD, Activity.Description + account  ~ variable, c(sum,mean,min,max), margins = TRUE),scientific = FALSE)
```
##Can you make a bar plot or a histogram? Can you compare two related quantities?
Yes, both can be accomplished below we will compare yearly results to budget in Contracted Services and Materials and Services for the three years sampled.
```{r}
#Plot 2015 Contracted Services: subset data
CSPlot2015 <- filter(GL_Final,grepl("2015", posting_yr) &  grepl("O&M", Designation) &  grepl("62",expense_code))


CSBar15 <- ggplot(CSPlot2015, aes(posting_pd, SumOftrans_amt))+
         geom_bar(stat = "identity", aes(fill = Types), position = "dodge", na.rm = TRUE)+
  xlab("posting_pd")+ylab("Expenditures")+theme_bw()
        
CSBar15  

#Plot 2016 Contracted Services: subset data
CSPlot2016 <- filter(GL_Final,grepl("2016", posting_yr) &  grepl("O&M", Designation) &  grepl("62",expense_code))


CSBar16 <- ggplot(CSPlot2016, aes(posting_pd, SumOftrans_amt))+
         geom_bar(stat = "identity", aes(fill = Types), position = "dodge", na.rm = TRUE)+
  xlab("posting_pd")+ylab("Expenditures")+theme_bw()
        
CSBar16

#Plot 2017 Contracted Services: subset data
CSPlot2017 <- filter(GL_Final,grepl("2017", posting_yr) &  grepl("O&M", Designation) &  grepl("62",expense_code))


CSBar17 <- ggplot(CSPlot2017, aes(posting_pd, SumOftrans_amt))+
         geom_bar(stat = "identity", aes(fill = Types), position = "dodge", na.rm = TRUE)+
  xlab("posting_pd")+ylab("Expenditures")+theme_bw()
        
CSBar17

#Plot 2015 Materials & Services: subset data
MSPlot2015 <- filter(GL_Final,grepl("2015", posting_yr) &  grepl("O&M", Designation) &  grepl("47",expense_code))


MSBar15 <- ggplot(MSPlot2015, aes(posting_pd, SumOftrans_amt))+
         geom_bar(stat = "identity", aes(fill = Types), position = "dodge", na.rm = TRUE)+
  xlab("posting_pd")+ylab("Expenditures")+theme_bw()
        
MSBar15  

#Plot 2016 Materials & Services: subset data
MSPlot2016 <- filter(GL_Final,grepl("2016", posting_yr) &  grepl("O&M", Designation) &  grepl("47",expense_code))


MSBar16 <- ggplot(MSPlot2016, aes(posting_pd, SumOftrans_amt))+
         geom_bar(stat = "identity", aes(fill = Types), position = "dodge", na.rm = TRUE)+
  xlab("posting_pd")+ylab("Expenditures")+theme_bw()
        
MSBar16

#Plot 2017 Materials & Services: subset data
MSPlot2017 <- filter(GL_Final,grepl("2017", posting_yr) &  grepl("O&M", Designation) &  grepl("47",expense_code))


MSBar17 <- ggplot(MSPlot2017, aes(posting_pd, SumOftrans_amt))+
         geom_bar(stat = "identity", aes(fill = Types), position = "dodge", na.rm = TRUE)+
  xlab("posting_pd")+ylab("Expenditures")+theme_bw()
        
MSBar17
```

##Can you make a scatterplot? Can you make a time-series plot?
Yes we will scatterplot expenditures by month for each year sampled for Contracted Services & Materials & Supplies.  This is both a timeseries and a scatter plot.  
```{r, scatterplot}
#Contracted Services 2015
ggplot(CSPlot2015, aes(x = posting_pd, y = SumOftrans_amt, color = activity )) +
  geom_point(data = subset(CSPlot2015, Types == "ACTUAL"))+
  scale_color_gradient()+
  ggtitle("2015 Contracted Services by Activity Code")+
  labs(x = "Posting Period", y = "Expenditures")

#Contracted Services 2016
ggplot(CSPlot2016, aes(x = posting_pd, y = SumOftrans_amt, color = activity )) +
  geom_point(data = subset(CSPlot2015, Types == "ACTUAL"))+
  scale_color_gradient()+
  ggtitle("2016 Contracted Services by Activity Code")+
  labs(x = "Posting Period", y = "Expenditures")

#Contracted Services 2017
ggplot(CSPlot2017, aes(x = posting_pd, y = SumOftrans_amt, color = activity )) +
  geom_point(data = subset(CSPlot2015, Types == "ACTUAL"))+
  scale_color_gradient()+
  ggtitle("2017 Contracted Services by Activity Code")+
  labs(x = "Posting Period", y = "Expenditures")
#Materials & Supplies 2015
ggplot(MSPlot2015, aes(x = posting_pd, y = SumOftrans_amt, color = activity )) +
  geom_point(data = subset(MSPlot2015, Types == "ACTUAL"))+
  scale_color_gradient()+
  ggtitle("2015 Materials & Supplies by Activity Code")+
  labs(x = "Posting Period", y = "Expenditures")

#Materials & Supplies 2016
ggplot(MSPlot2016, aes(x = posting_pd, y = SumOftrans_amt, color = activity )) +
  geom_point(data = subset(MSPlot2015, Types == "ACTUAL"))+
  scale_color_gradient()+
  ggtitle("2016 Materials & Supplies by Activity Code")+
  labs(x = "Posting Period", y = "Expenditures")

#Materials & Supplies 2017
ggplot(MSPlot2017, aes(x = posting_pd, y = SumOftrans_amt, color = activity )) +
  geom_point(data = subset(MSPlot2015, Types == "ACTUAL"))+
  scale_color_gradient()+
  ggtitle("2017 Materials & Supplies by Activity Code")+
  labs(x = "Posting Period", y = "Expenditures")


  
```
##Having made these plots, what are some insights you get from them? Do you see any correlations? Is there a hypothesis you would like to investigate further? What other questions do they lead you to ask?

These plots have shown that to our departments, annual budget is more important than getting the monthly allocations reflected accordingly. This furthers the need for accurate forecasting for budget development. At this point in time, it is apparent when there are outage expendiutres, the lowest of activity codes, and that there are spikes in these areas. It appears that seasonal breakdown of expenditutres will be a good way of developing a forecasting method. 




